Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar

### 2019-01-08 accessPEP.Rmd
adding in some code to show patient daily ts
## don't know if it's worth the pain!

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  html_document: default
  pdf_document:
    includes:
      in_header: mystyle.sty
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
css: style.css
---
## Goals
- To estimate burden of rabies in Madagascar using nationally available data from clinics that provide PEP.
- To explore the potential impacts of expanded access to PEP

## Data
- Patient data from 19 ARMC accross the country over four years (2013 - 2017)
  - includes dates and home locations (to the district level) of patient reporting for PEP
- Patient data from Moramanga ARMC over 28 months (Sep 2016 - Dec 2018)
  - includes dates and home locations (to the commune level) of patients reporting for PEP
  - also includes rabies exposure status based on a combination of clinic-based triage and contact
  tracing (Rajeev et al. 2018)
- Travel times estimated from the Malaria Atlas Project's global friction surface for 2015
  - We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al. 2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale. 
  - We then calculated a weighted average by human population to the commune or district level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs.
  - For each clinic, we defined the catchment area as all communes for which the clinic was the closest ARMC.
- Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the designated admin level.

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
```

@@

@@

## Get color palette from other network fig!
img <- readJPEG("~/Downloads/Dp0j4GNUwAAPPJV.jpeg")
pal <- image_palette(img, 31)
add.alpha <- function(col, alpha=1){
  if(missing(col))
    stop("Please provide a vector of colours.")
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                       rgb(x[1], x[2], x[3], alpha=alpha))  
}
ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
                 "#63FFAC", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
ctar_metadata$fill <- ctar_metadata$color
ctar_metadata$fill[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.25)

display_palette(ctar_metadata$color)
dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
mada_district@data <- left_join(mada_district@data, dist_catchments)

## get exposure matrix
ctar_data %>% 
  group_by(district, id_ctar) %>% # group by district and ctar
  summarize(n = n()) -> exp_mat # get count of exposures
dist_points <- cbind(select(mada_district@data,
                            mdg_dis_co), coordinates(mada_district))
names(dist_points)[2:3] <- c("dist_long", "dist_lat")
dist_points %>%
 left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
           by = c("mdg_dis_co" = "ctar_dist")) %>%
 mutate(dist_long = coalesce(dist_long, LONGITUDE), 
        dist_lat = coalesce(dist_lat, LATITUDE)) -> dist_points

exp_mat %>% 
  left_join(select(dist_points, mdg_dis_co, dist_long, dist_lat), 
            by = c("district" = "mdg_dis_co")) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines$size <- log(dist_lines$n + 0.1)*0.2

ctar_data %>%
  group_by(district) %>%
  summarize(n = n()) %>%
  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
             by = c("district" = "mdg_dis_co")) %>%
  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
dist_points$size <- log(dist_points$n + 0.1)*0.25

pdf("figs/network_1.pdf")
plot(mada_district, col = mada_district$fill, border = "grey") 
points(jitter(dist_lines$dist_long, factor = 2), jitter(dist_lines$dist_lat, factor = 2)
       , pch = 1, cex = dist_points$size, 
       col = dist_lines$color)
segments(dist_lines$LONGITUDE, dist_lines$LATITUDE, x1 = dist_lines$dist_long, y1 = dist_lines$dist_lat,
         lwd = dist_lines$size*2, col = dist_lines$color)
dev.off()

exp_mat %>% 
  left_join(select(ctar_metadata, CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, CTAR, exclude), by = c("district" = "mdg_dis_co")) %>%
  group_by(district) %>%
  mutate(prop = n/sum(n),
         catch_match = ifelse(CTAR.x == CTAR.y, 1, 0)) -> catch_check
hist(catch_check$prop)
ggplot(data = catch_check, aes(CTAR.x, district)) + 
  geom_tile(aes(fill = prop)) +
  scale_fill_continuous(type = "viridis", name = "proportion") +
  geom_text(label = catch_check$catch_match)
```

```{r clinic level reporting}
######### Patient time series
ctar_data %>%
  select(id_ctar, date_de_consultation) %>%
  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
  gather(dose, date_de_consultation, -id_ctar) %>%
  group_by(id_ctar, date_de_consultation) %>%
  summarise(n = n()) -> patient_ts
patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]

ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
  geom_tile(aes(fill = n)) +
  scale_fill_continuous(type = "viridis", name = "Patients")

## getting reporting
start_date <- ymd("2014-01-01")
end_date <- ymd("2017-12-31")
ts <- as_tibble(seq(start_date, end_date, by = "day"))

@@

date_mat <- matrix(NA, nrow(dose_mat), ncol(dose_mat))
for (j in 1:ncol(date_mat)){
  rle(dose_mat[ , j]) %>%
    unclass() %>%
    as.data.frame() %>%
    mutate(end = cumsum(lengths),
         start = c(1, lag(end)[-1] + 1)) %>%
    filter(values == 0, lengths >= 10) -> rles
  for (i in 1:nrow(rles)){
    date_mat[rles$start[i]:rles$end[i], j] <- 0
  }
}

date_mat <- replace(date_mat, is.na(date_mat), 1)
date_mat %>% 
  as_tibble() %>%
  group_by(date = year(doses_wide$date_de_consultation)) %>%
  summarise_all(funs(sum(.)/n())) -> clinic_reporting
colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]
clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]
clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)

ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")
```

to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
ctar_data$excl <- 0
ctar_data$excl[ctar_data$remarque %in% to_exclude] <- 1

ctar_data %>% 
  filter(excl != 1) %>% # exclude contacts
  mutate_at(vars(starts_with("date")), funs(ymd(.))) %>% # format dates
  group_by(year = year(date_de_consultation), district, id_ctar) %>% # group by year and district
  summarize(n = n()) -> exposures # get count of exposures

exposures %>%
  rename(date = year) %>%
  left_join(covars) %>% # add pop + ttimes + reporting
  left_join(clinic_reporting) %>%   # add reporting by clinic + year
  filter(exclude == 0, prop > 0.25) -> exps_cleaned

exps_cleaned %>%
  group_by(district, date) %>%
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) %>%
  group_by(district) %>%
  summarise(bites = mean(bites)) %>%
  left_join(covars) -> exps_dist

exps_cleaned %>%
  group_by(ctar, date) %>% ## Need to rename CTAR so clear if it's CTAR in catchment vs. ctar of bites...
  mutate(bites = n/prop) %>%
  summarise(bites = sum(bites, na.rm = TRUE)) %>%
  group_by(ctar) %>%
  summarise(bites = mean(bites)) -> exps_ctar

@@

```{r likelihood of observing district data given sum of communes in district}
library()

get.likelihood <- function(commune,district, params) {
  # To test likelihood function
  # commune = comm_data
  # district = exps_dist
  # params = c(1e-6, 0.1)
  B_ttimes <- params[1]
  B_0 <- params[2]
  commune %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(district, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
  return(-sum(dpois(round(sum_bites$dist_bites), lambda = sum_bites$sum_bites, log = TRUE)))
}

tmp <- optim(par = c(1e-6, 0.1), get.likelihood, district = exps_dist,
             commune = comm_data, hessian = TRUE)
params <- tmp$par
B_ttimes <- params[1]
B_0 <- params[2]

comm_data %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) -> check
plot(check$ttimes_weighted, check$bites)

comm_data %>%
      filter(exclude == 0) %>%
      group_by(mdg_dis_co) %>%
      mutate(bites = exp(B_ttimes*ttimes_weighted + B_0)*pop) %>%
      summarize(sum_bites = sum(bites)) %>%
      left_join(select(district, district, dist_bites = bites), 
                by = c("mdg_dis_co" = "district")) %>%
      filter(!is.na(dist_bites)) -> sum_bites
sum_bites %$% plot(dist_bites, sum_bites)
abline(a = 0, b = 1, col = "blue", lty = 2)

@@

## Model of reported bites as a function of travel time
As bite patient data were only available...

Excluding any catchments for which there was less than 25% estimated form submission, we calculated the average bite incidence corrected by the annual undersubmission of forms at the district level.

We fit a linear mixed effects model where the annual average bite incidence at the district level was predicted by the average travel time to the nearest CTAR for the district, the population within the district, and a random catchment level effect.

For three other catchments with reporting above x% we used to out-of-fit predict...
Ability to predict average annual/monthly bite incidence by travel time
Ability to predict out-of-fit to other catchments

We calculated deaths as:
$$deaths_{i} = (R_{i} - ({p_{rabid} \times B_{i})) \times p_{death}$$

>>>>

---
title: "Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar"
author: ""
date: 
output:
  html_document: default
  pdf_document:
    includes:
      in_header: mystyle.sty
  word_document:
    reference_docx: format_docs/word_styles_reference_01.docx
css: style.css
---

## INTRODUCTION

Access to healthcare is a significant driver of disease burden globally. Spatial access to care (cite Jess paper here). Rarely incorporated into national estimates of burden due to lack of data and tools for understanding...

Canine rabies causes an estimated 59,000 human deaths annually. Post-exposure prophylaxis (PEP) prevents fatalities if delivered in a timely manner, however, access to this intervention is highly limited in areas where the disease is endemic. Data on true rabies exposures in humans and incidence in animals is also lacking in most of these countries. Often, the only data available are numbers of bite victims reporting to health facilities. The majority of rabies burden studies use these data to estimate burden of rabies from probability decision tree frameworks, often with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more overall bites reported in a location, the higher the incidence of *rabies* exposures there) and that reporting is uniform across space.

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to the government or to patients that present to 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. Given that rabies is endemic in the country and there is effectively no vaccination of domestic dogs, we use the baseline predictions of incidence from the dynamic model reported in Hampson et al. 2018. We flip the standard decision tree and make the assumption that reported bite incidence reflects access and reporting to PEP rather than differences in rabies incidence, using travel times to clinics as a predictor of bites reported to clinics. Assuming a range of exposure incidence given endemic rabies, we generate sub-national estimates of rabies burden and explore the impacts of expanding access to PEP in Madagascar.

The most recent estimation of burden and the impact of PEP used another approach, using transmission dynamic models as the backbone to predict incidence based on level of vaccination coverage and size of the dog population at the national level. Using transmission dynamic models to estimate incidence could improve upon previous studies which may underestimate rabies burden in areas with low reporting. Here, we adapt this framework and use bite patient data from Madagascar to predict rabies burden.

## METHODS

### Bite patient data
We used the database of bite patient forms submitted from 24 ARMC across ARMC between 2014 - 2017. These were individual patient data forms that were submitted to IPM as frequently as monthly to annually by clinics. Two clinics, the IPM ARMC and the Fianarantsoa ARMC used computer databases from which the data during this period were extracted and merged to the larger database. These data include the administrative district of the bite patient's address. As previous analyses showed likely substantial undersubmission of forms from clinics (*cite baseline data paper*), we calculated the proportion of days of the year for which forms were submitted. We excluded periods with 10 or more consecutive dates with no patient forms and calculated the proportion of days included per year.

- Patient data from 19 ARMC accross the country over four years (2013 - 2017)

  - includes dates and home locations (to the district level) of patient reporting for PEP
- Patient data from Moramanga ARMC over 28 months (Sep 2016 - Dec 2018)
  - includes dates and home locations (to the commune level) of patients reporting for PEP

RIG data?

### GIS Data
We use the global friction surface for 2015 generated by the Malaria Atlas Project ( https://map.ox.ac.uk/research-project/accessibility_to_cities/, Weiss et al. 2015,) and GPS points of clinics to get travel time to the nearest ARMC for the country of Madagascar at a 1 x 1 km scale. We then calculated a weighted average of travel times by human population to the commune level, using administrative shapefiles available trhough the UN Office for the Coordination of Humanitarian Affairs. For each clinic, we defined the catchment area as all **districts?** for which the clinic was the closest ARMC. Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, Linaird et al. 2012) and also aggreagated to the commune level
<br>

### Model of reported bites as a function of travel time
While the national bite patient data is available at the district level, travel times vary significantly within a district (see Figure SX). In order to translate the impacts of differences in access at sub-district scales to the magnitude of reported bites at the district scale, we modeled bites at the district level as the sum of incidence at the commune level. Incidence at the commune level is then a function of travel times to the closest ARMC. Specifically, we modeled bites as follows:
$$ 
  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)\times pop_j 
$$
where $\mu_{d}$ is the mean number of bites in district, which is the sum of bites at the commune level given **commune level travel times**, $T_j$. We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.

As we had data available on reported bites at the commune level from the Moramanga ARMC, we modeled observed commune bites in the same framework, but un-aggregated where:
$$ 
\mu_{j} = exp(\beta_tT_j + \beta_0)\times pop_j 
$$
where $\mu_{j}$ = the mean number of bites in commune $j$ and the observed bites at the commune level follow a poisson distribution around the mean $\mu_d$. We only included communes which were designated to be within the catchment for the clinic (so primarily the Moramanga and Anosibe A'nala Districts)?? We compared our estimates of $\beta_t$ (i.e. the impact of travel time on incidence) and $\beta_0$ (the intercept) for our district data at the national level and the commune level data from the Moramanga ARMC.

To better understand if ttimes might capture access, we also looked at whether pop alone can predict bites and distance could predict bites...
Ability to predict out-of-fit to other catchments? Tana + Fianarantsoa?

We calculated deaths as:
$$
deaths_i = (R_i - p_{rabid}B_i) \times p_{death}
$$

where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
$$
           p_{rabid}=
            \begin{cases}
            x, & \text{if}\ \frac{R_i \times \rho_{max}}{B_i} > x \\
            \frac{R_i \times \rho_{max}}{B_i}, & \text{otherwise}
            \end{cases}
$$
<br>

such that rabid reported bites (i.e. $p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e. $R_i \times \rho_{max}$). $\rho_{max}$ taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e. the area with the minimum travel time). We look at the range of $x$ reported in Rajeev et al. 2018 for data from the Moramanga District (0.2 - 0.6).

$R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures. We calculate this as:

where $r$ is rabies incidence in dogs in the absence of any vaccination multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al. 2018). We estimate the dog population by using a human:dog ratio of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence. As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).

### Sensitivity Analyses
- R (increasing or decreasing with pop) (min + max)
- P-rabid min and max
- Rho-max min and max
- Probability of death min and max

### Estimating the impact of expanding PEP Access
We use this framework to compare three scenarios of PEP provisioning in Madagascar:

1. The baseline with the current clinic locations (n = 31)
2. Expansion of ARMC to one clinic per district (n = 114)
3. Expansion of ARMC to all CSB IIs, something to describe CSB IIs here, that is primary hospitals (n = X)

## RESULTS

```{r directory, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
rm(list = ls())

## Libraries
library(knitr)
library(RImagePalette)
library(jpeg)
library(tidyverse)
library(lubridate)
library(magrittr)
library(rgdal)
library(broom)
library(stringr)
library(ISOweek)
library(raster)
library(sp)

source("R/utils.R")

select <- dplyr::select
knitr::opts_chunk$set(fig.pos = 'h', echo = FALSE, message = FALSE, warning = FALSE, results = 'hide')
```

@@

### Clinic catchments and reporting
For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1B), although some patients did report to the ARMC that were not closest to them in terms of travel times (Figure 1A). In addition, at the clinic level, the majority of bites reported to a given ARMC came from within the catchment (Figure 1C).

@@

## Get color palette from other network fig!
ctar_metadata$color <- c("#FCC56F","#FFDBE5", "#7A4900", "#CBCDD2", "#0000A6", "#EFF2F1",
                 "#99d8c9", "#B79762", "#004D43", "#8FB0FF", "#997D87", "#FD9C54", "#8362B5",
                 "#578FB0","#5A0007", "#809693", "#D16100", "#1B4400", "#4FC601", "#3B5DFF", 
                 "#4A3B53", "#FF2F80","#61615A", "#BA0900", "#6B7900", "#00C2A0", "#FFAA92",
                 "#FF90C9", "#B903AA", "#FEFFE6", "#E9E9D2")
ctar_metadata$color[ctar_metadata$CTAR %in% no_data] <- "#D3D3D3"
ctar_metadata$fill <- ctar_metadata$color
ctar_metadata$fill <- add.alpha(ctar_metadata$fill, 0.4)
display_palette(ctar_metadata$color)

dist_catchments %>% left_join(select(ctar_metadata, CTAR, exclude, 
                                     color, fill, ctar_dist, LATITUDE, LONGITUDE)) -> dist_catchments
mada_district@data <- left_join(mada_district@data, dist_catchments) ## need this?

## get exposure matrix = first set of points
## district exposures
ctar_data %>% 
  drop_na(id_ctar, district) %>%
  group_by(district) %>% # group by district
  summarize(n = n()) -> exps_dist # get count of exposures
## ctar exposures
ctar_data %>% 
  drop_na(id_ctar) %>%
  group_by(id_ctar) %>% # group by ctar
  summarize(n = n()) -> exps_ctar
## ctar by district exposure matrix
ctar_data %>% 
  drop_na(id_ctar, district) %>%
  group_by(district, id_ctar) %>% # group by district and ctar
  summarize(n = n()) -> exp_mat# get count of exposures

## To plot district bites
dist_points <- cbind(select(mada_district@data,
                            mdg_dis_co), coordinates(mada_district))
names(dist_points)[2:3] <- c("dist_long", "dist_lat")
dist_points %>%
 left_join(select(ctar_metadata, ctar_dist, LATITUDE, LONGITUDE), 
           by = c("mdg_dis_co" = "ctar_dist")) %>%
 mutate(dist_long = coalesce(LONGITUDE, dist_long), 
        dist_lat = coalesce(LATITUDE, dist_lat)) -> dist_points

exps_dist %>%
  right_join(select(dist_points, mdg_dis_co, dist_long, dist_lat),
             by = c("district" = "mdg_dis_co")) %>%
  left_join(dist_catchments, by = c("district" = "mdg_dis_co") ) -> dist_points
dist_points$size <- log(dist_points$n + 0.1)*0.25

## To plot ctar points (only those not excluded)
exps_ctar %>%
 left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE, 
                  exclude, color, fill)) %>%
 filter(exclude == 0) -> ctar_points

## To plot lines between ctar + districts
exp_mat %>% 
  left_join(select(dist_points, district, dist_long, dist_lat)) %>%
  left_join(select(ctar_metadata, id_ctar, LATITUDE, LONGITUDE,
                   color, fill, ctar_dist)) -> dist_lines
dist_lines$size <- log(dist_lines$n + 0.1)*0.2

pdf("figs/network.pdf")
plot(mada_district, col = mada_district$fill, border = "grey") 
segments(dist_lines$dist_long, dist_lines$dist_lat, x1 = dist_lines$LONGITUDE, 
         y1 = dist_lines$LATITUDE,
         lwd = dist_lines$size*2, col = dist_lines$color)
points(ctar_points$LONGITUDE, ctar_points$LATITUDE, col = ctar_points$color, pch = 1,
       cex = log(ctar_points$n + 0.1)*0.25)
points(dist_points$dist_long, dist_points$dist_lat, cex = dist_points$size, pch = 20,
       col = dist_points$color)
dev.off()

exp_mat %>% 
  left_join(select(ctar_metadata, ctar_obs = CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, ctar_catch = CTAR, exclude), 
            by = c("district" = "mdg_dis_co")) %>%
  group_by(district) %>% # group by ctar
  mutate(prop = n/sum(n),
         catch_match = ifelse(ctar_obs == ctar_catch, 1, 0)) %>%
  filter(catch_match == 1) -> catch_dist
hist(catch_dist$prop, main = "", 
     xlab = "Proportion of patients \n reporting to closest ARMC",
     ylab = "Number of districts", col = "grey50", border = "white", xlim = c(0, 1))

exp_mat %>% 
  left_join(select(ctar_metadata, ctar_obs = CTAR, id_ctar)) %>%
  left_join(select(dist_catchments, mdg_dis_co, ctar_catch = CTAR, exclude), 
            by = c("district" = "mdg_dis_co")) %>%
  group_by(ctar_obs) %>% # group by ctar
  mutate(prop = n/sum(n),
         catch_match = ifelse(ctar_obs == ctar_catch, 1, 0)) %>%
  filter(catch_match == 1) %>%
  summarize(prop = sum(prop)) -> catch_ctar
hist(catch_ctar$prop, main = "", 
     xlab = "Proportion of patients reporting from within catchment",
     ylab = "Number of ARMC", col = "grey50", border = "white", xlim = c(0, 1))
```

We estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 10 consecutive days (Figure 2). We further excluded any years for which there was less than 25% reporting.

```{r clinic level reporting}
######### Patient time series
ctar_data %>%
  select(id_ctar, date_de_consultation) %>%
  drop_na(id_ctar) %>%
  mutate(date_de_consultation = ymd(date_de_consultation)) %>%
  gather(dose, date_de_consultation, -id_ctar) %>%
  group_by(id_ctar, date_de_consultation) %>%
  summarise(n = n()) -> patient_ts

patient_ts <- patient_ts[!is.na(patient_ts$date_de_consultation), ]
patient_ts$ctar <- ctar_metadata$CTAR[match(patient_ts$id_ctar, ctar_metadata$id_ctar)]
patient_ts <- patient_ts[!is.na(patient_ts$ctar), ]

start_date <- ymd("2014-01-01")
end_date <- ymd("2017-12-31")

## getting reporting
ts <- as_tibble(seq(start_date, end_date, by = "day"))

@@

get.days <- function(dmat = dose_mat, threshold = 10) {
  date_mat <- matrix(NA, nrow(dmat), ncol(dmat))
  for (j in 1:ncol(date_mat)){
    rle(dmat[ , j]) %>%
      unclass() %>%
      as.data.frame() %>%
      mutate(end = cumsum(lengths),
             start = c(1, lag(end)[-1] + 1)) %>%
      filter(values == 0, lengths >= threshold) -> rles
  if(nrow(rles) > 0){
    for (i in 1:nrow(rles)){
        date_mat[rles$start[i]:rles$end[i], j] <- 0
      }
    }
  }

  date_mat <- replace(date_mat, is.na(date_mat), 1)
  date_mat %>% 
    as_tibble() %>%
    group_by(date = year(doses_wide$date_de_consultation)) %>%
    summarise_all(funs(sum(.)/n())) -> clinic_reporting
  colnames(clinic_reporting) <- colnames(doses_wide)[1:(ncol(doses_wide)-1)]

  clinic_reporting <- gather(clinic_reporting, id_ctar, prop, -date_de_consultation)
  clinic_reporting$ctar <- ctar_metadata$CTAR[match(clinic_reporting$id_ctar, 
                                                  ctar_metadata$id_ctar)]
  clinic_reporting$id_ctar <- as.numeric(clinic_reporting$id_ctar)
  clinic_reporting$threshold <- threshold
  return(list(clinic_reporting, date_mat))
}

## Color or size dots by their # from that year
# ctar_data %>%
#   group_by(date_de_consultation = year(date_de_consultation), id_ctar) %>%
#   summarize(tot = n()) -> year_by_clinic

clinic_reporting <- get.days(threshold = 10)[[1]]
# clinic_reporting %>%
#   mutate(prop10 = prop, prop5 = get.days(threshold = 5)[[1]]$prop, 
#          prop30 = get.days(threshold = 30)[[1]]$prop) %>%
#   left_join(year_by_clinic) %>%
#   ggplot(., aes(x = reorder(ctar, prop10), y = prop10)) + 
#   geom_point() +
#   geom_linerange(aes(ymin = prop30, ymax = prop5)) +
#   facet_wrap(~ date_de_consultation, nrow = 1) +
#   coord_flip() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   xlab("") + 
#   ylab("Proportion of days included") +
#   scale_color_gradient(low = "white", high = "blue")
# 
# date_mat <- as.data.frame(get.days(threshold = 10)[[2]])
# date_mat$date_de_consultation <- doses_wide$date_de_consultation
# names(date_mat[1:(ncol(date_mat)-1)]) <- names(doses_wide[2:(ncol(date_mat))])
# library(reshape2)
# head(melt(date_mat, id = "date_de_consultation"))

ggplot(data = patient_ts, aes(date_de_consultation, ctar)) + 
   geom_tile(aes(fill = n)) +
   scale_fill_continuous(type = "viridis", name = "Patients") +
   xlim(start_date, end_date)

ggplot(clinic_reporting, aes(x = date_de_consultation, y = ctar, fill = prop)) + 
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue")

```{r getting Mora data}
patient_ts %>%
  group_by(ctar) %>%
  mutate(exclude = ifelse(n > mean(n) + 6*sd(n), 1, 0)) -> patient_ts

ggplot(data = drop_na(patient_ts, exclude), aes(x = ctar, y = n, color = as.factor(exclude))) +
  geom_jitter(alpha = 0.5, width = 0.25) +
  coord_flip()

to_exclude <- c(grep("contact", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("contam", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("consom", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE),
                grep("passage", unique(ctar_data$remarque), ignore.case = TRUE, value = TRUE))
to_exclude <- to_exclude[-grep("date", to_exclude, ignore.case = TRUE)]
ctar_data$exclude_bycomment <- 0
ctar_data$exclude_bycomment[ctar_data$remarque %in% to_exclude] <- 1
ctar_data <- mutate(ctar_data, date_de_consultation = ymd(date_de_consultation))
patient_ts %>%
  ungroup() %>%
  select(id_ctar, date_de_consultation, exclude_bydate = exclude) %>%
  right_join(ctar_data) -> ctar_check

(ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/
  length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])

ctar_check %>%
  group_by(id_ctar) %>%
  summarize(prop_excluded = length(exclude_bydate[exclude_bydate == 1])/n()) %>%
  ggplot(data = ., aes(y = id_ctar, prop_excluded)) +
  geom_point()

@@

``` {r check with moramanga}
## ctar for # bites and also checking days to rule out contacts
col_lookup <- read.csv("data/new_colnames.csv", header = TRUE)
ctar <- read.csv("data/moramanga/ctar_up_todate.csv", header = TRUE)
ctar <- match.colnames(ctar, subset(col_lookup, form == "ctar"))

## merge with newest data!

## Set-up time series
tfunc <- function (date, format_date = "%Y-%m-%d", start = "01-09-2016", format_start = "%d-%m-%Y",
                   year1 = 2016, tstep = "month", period = FALSE) {
  get.consec (date, format_date, start, format_start,
              year1, tstep, period) - 8 ## starts in Sep
}

## all dates
ctar %>%
  mutate_at(vars(starts_with("date")), funs(dmy(.))) -> ctar
ctar$month <- tfunc(ctar$date.reported)
## subset to study period (Sep 2016 - to date)
## checks for cases where month is NA!
ctar <- subset(ctar, month > 0 & month < 28)

# ctar district and commune
ctar$commune <- sapply(strsplit(as.character(ctar$patient.location), "\\("), "[", 1)
ctar$commune <- trimws(ctar$commune, which = "right")
ctar$district <- sapply(strsplit(as.character(ctar$patient.location), "\\, "), "[", 2)
ctar$district <- gsub(" \\(District\\)", "", ctar$district)

## change all true/false to 0/1
ctar[ctar == FALSE] <- 0

## Bite data in Moramanga
n_months <- max(ctar$month, na.rm = TRUE)

ctar %>%
  left_join(select(mada_communes@data, commune, mdg_cm_, mtch_ds), 
            by = c("commune" = "commune", "district" = "mtch_ds")) %>%
  filter(consult.type == "Animal bite/exposure") %>%
  group_by(mdg_cm_) %>%
  summarize(bites = round(n()/n_months*12)) %>%
  left_join(comm_covars) %>%
  filter(!is.na(ttimes_weighted), 
         district %in% mada_district@data$district[mada_district@data$CTAR == "Moramanga"]) -> mora_bites

ctar %>%
  group_by(date.reported) %>%
  summarize(n = n()) %>%
  mutate(exclude = ifelse(n > mean(n) + 2*sd(n), 1, 0)) -> mora_ts

ggplot(data = mora_ts, aes(x = "Moramanga", y = n, color = as.factor(exclude))) +
  geom_jitter(alpha = 0.5, width = 0.1) +
  coord_flip()

ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
table(ctar$consult.type, ctar$exclude)

sdev <- seq(2, 10, by = 1)
check <- data.frame(sd = NA, prop_bites = NA, prop_contacts = NA)
for (i in 1:length(sdev)){
  ctar %>%
    group_by(date.reported) %>%
    summarize(n = n()) %>%
    mutate(exclude = ifelse(n > mean(n) + sdev[i]*sd(n), 1, 0)) -> mora_ts
  ctar$exclude <-mora_ts$exclude[match(ctar$date.reported, mora_ts$date.reported)]
  checksd <- table(ctar$consult.type, ctar$exclude)[, 2]/table(ctar$consult.type)
  check <- rbind(check, c(sdev[i], checksd[1:2]))
}

plot(check$sd, check$prop_contacts, col = "blue", pch = 20, bty = "l", ylim = c(0,1))
points(check$sd, check$prop_bites, col = "red", pch = 20)

@@

### Models of bites as a function of travel times
